A Logical Solution of the Linearity Challenge

### 2025-12-02 paper/Congruence.lagda.tex
revised congruence

The constructor \AgdaInductiveConstructor{s-comm} models the commutativity
property of parallel composition. We use \AgdaFunction{dual-symm} and
\AgdaFunction{+-comm} to compute the proofs of the relation
$\AgdaDatatype{Dual}~B~A$ and of the splitting $\ContextC \simeq \Context_2 +
\Context_1$ from $d$ and $p$ respectively.

implicit rewriting

@@

There are two non-trivial aspects worth commenting. The first one concerns the
proof $(\AgdaInductiveConstructor{split-r}~q)$ used in the
\AgdaInductiveConstructor{wait} process before structural precongruence is
applied. To understand the meaning of this proof, we must recall three key
elements:
\begin{enumerate}
  \item
    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{split-r}~q)~P)$
    is a direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore
    it is meant to be well typed in the context
    $A~\AgdaInductiveConstructor{∷}~\Context_1$.
  \item Being a \AgdaInductiveConstructor{wait} process, such context must
    contain a $\Bot$ type as per the typing rule \WaitRule. That is
    $A~\AgdaInductiveConstructor{∷}~\Context_1 \simeq [\Bot] +
    A~\AgdaInductiveConstructor{∷}~\Delta$ for some $\Delta$.
  \item The \SWait rule is applicable only provided that the channel restricted
    by the cut (say $x$, of type $A$) is different from the channel (say $y$, of
    type $\Bot$) consumed by the \AgdaInductiveConstructor{wait} process.
    %
    We enforce the side condition $x \ne y$ of \SWait imposing that the type $A$
    in front of $A~\AgdaInductiveConstructor{∷}~\Context_1$ necessarily comes
    from the right component of the splitting $[\Bot] +
    A~\AgdaInductiveConstructor{∷}~\Delta$ through the use of
    \AgdaInductiveConstructor{split-r}.
\end{enumerate}

@@

The constructors \AgdaInductiveConstructor{s-select-l} and
\AgdaInductiveConstructor{s-select-r} model \SSelect when the selected tag is
respectively \AgdaInductiveConstructor{true} and
\AgdaInductiveConstructor{false}.

@@

Here the process
$(\AgdaInductiveConstructor{select}~\AgdaInductiveConstructor{true}~(\AgdaInductiveConstructor{split-r}~q)~P)$,
which is found under a cut for $x : A$, is using some channel $y : C \Plus D$ to
send the tag $\AgdaInductiveConstructor{true}$.
%
Unlike \AgdaInductiveConstructor{wait}, which consumes the channel it operates
on, the continuation process $P$ will be using a fresh continuation channel $z :
C$. Therefore, $P$ is required to be well typed in the context
$C~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
the type of the continuation sits on top of the type of the restricted channel.
%
After structural precongruence is applied, however, the type $C$ of the
continuation channel $z$ ends up underneath that of the restricted channel $x$,
because now the two channels are introduced in the opposite order. Therefore, we
are in a situation where we need to turn $P$ into a process that is well typed
in the context
$A~\AgdaInductiveConstructor{∷}~C~\AgdaInductiveConstructor{∷}~\ContextD$. This
is achieved applying the function \AgdaFunction{\#process} to the
\AgdaInductiveConstructor{\#here} permutation, which swaps the two topmost types
in a typing context, and finally to the process $P$.

\emph{En passant}, we also show the modeling of the \SForkL rule, which is
interesting because of its complex side conditions:

@@

Finally, in \Cref{sec:semantics} we have colloquially defined $\pcong$ as a
``\emph{pre-congruence}'', implicitly implying that it is reflexive, transitive,
and that it is is preserved by some forms of calculus. In the formalization we
have to be precise and we introduce specific rules:

>>>>

The constructor \AgdaInductiveConstructor{s-comm} models the commutativity
property of parallel composition. We use \AgdaFunction{+-comm} to compute the
proof of the splitting $\Splitting\ContextC{\Context_2}{\Context_1}$ from $p$
respectively.
%
Notice that this constructor also makes key use of the implicit rewriting rule
\AgdaFunction{dual-inv} described in \Cref{sec:type-agda}. Indeed $P$ and $Q$
have type $\AgdaDatatype{Process}~(A~\AgdaInductiveConstructor{∷}~\Context_1)$
and
$\AgdaDatatype{Process}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\Context_2)$,
respectively, but the \AgdaInductiveConstructor{cut} on the r.h.s. of
\AgdaInductiveConstructor{⊒} expects $P$ to have type
$\AgdaFunction{dual}~(\AgdaFunction{dual}~A)$. Thanks to
\AgdaFunction{dual-inv}, Agda considers these types equivalent without requiring
intricate substitutions in the index of a data type.

@@

There are two non-trivial aspects worth commenting. The first one concerns the
proof $(\AgdaInductiveConstructor{>}~q)$ used in the
\AgdaInductiveConstructor{wait} process before structural precongruence is
applied. To understand the meaning of this proof, we must recall three key
elements:
\begin{enumerate}
  \item
    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{>}~q)~P)$ is a
    direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore it
    is meant to be well typed in the context
    $A~\AgdaInductiveConstructor{∷}~\Context_1$.
  \item Being a \AgdaInductiveConstructor{wait} process, such context must
    contain a $\Bot$ type as per the typing rule \WaitRule. That is
    $\Splitting{A~\AgdaInductiveConstructor{∷}~\Context_1}{[\Bot]}{A~\AgdaInductiveConstructor{∷}~\ContextD}$
    for some $\ContextD$.
  \item The \SWait rule is applicable only provided that the channel restricted
    by the cut (say $x$, of type $A$) is different from the channel (say $y$, of
    type $\Bot$) consumed by the \AgdaInductiveConstructor{wait} process.
    %
    We enforce the side condition $x \ne y$ of \SWait imposing that the type $A$
    in front of $A~\AgdaInductiveConstructor{∷}~\Context_1$ necessarily comes
    from the right component of the splitting $[\Bot] +
    A~\AgdaInductiveConstructor{∷}~\Delta$ through the use of
    \AgdaInductiveConstructor{>}.
\end{enumerate}

@@

The constructors \AgdaInductiveConstructor{s-left} and
\AgdaInductiveConstructor{s-right} model \SSelect when the selected tag is
respectively $\Tag_1$ and $\Tag_2$.

@@

Here the process
$(\AgdaInductiveConstructor{left}~(\AgdaInductiveConstructor{>}~q)~P)$, which is
found under a cut for $x : A$, is using some channel $y : B \Plus C$ to send the
tag $\Tag_1$.
%
Unlike \AgdaInductiveConstructor{wait}, which consumes the channel it operates
on, the continuation process $P$ will be using a fresh continuation channel $z :
B$. Therefore, $P$ is required to be well typed in the context
$B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
the type of the continuation sits on top of the type of the restricted channel.
%
After structural precongruence is applied, however, the type $B$ of the
continuation channel $z$ ends up underneath that of the restricted channel $x$,
because now the two channels are introduced in the opposite order. Therefore, we
are in a situation where we need to turn $P$ into a process that is well typed
in the context
$A~\AgdaInductiveConstructor{∷}~B~\AgdaInductiveConstructor{∷}~\ContextD$. This
is achieved applying the function \AgdaFunction{↭process} to the
\AgdaInductiveConstructor{swap} permutation, which swaps the two topmost types
in a typing context, and finally to the process $P$.

We also show the modeling of the \SForkL rule, which is interesting because of
its complex side conditions:

@@

Finally, in \Cref{sec:semantics} we have colloquially defined $\pcong$ as a
``\emph{pre-congruence}'', implying that it is reflexive, transitive relation
and that it is is preserved by some forms of calculus. In the formalization we
have to be precise and we introduce specific rules:
